Embedding Intuitionistic into Classical Logic

### 2023-03-09 LPAR/LPAR-EIICL.tex
Fixed small errors

\begin{definition}
For $\psi\in\mathcal X$ let $[\psi]$ denote the corresponding clause in $\mathcal X^{b S}$. Suppose we are given a counter-model $\mathcal M = (M, I)$ for $\varphi^{CH}$.
We say $(A\to B)\to C\in\mathcal X$ is \emph{fulfilled} at $m\in M$ iff $A^I(m)\to B^I(m)$ is false or $C^I(m)$ is true. For $\psi\in\mathcal X$ define $$g_\psi : M\to M, m\mapsto\begin{cases}
		m,&\text{ if $\psi$ is fulfilled at $m$,}\\
		f^I_{\lbrack\psi\rbrack}(m),&\text{ else.}		
	\end{cases}$$Define a model $\mathcal M_T = (M_T, I_T)$ as follows:
	\begin{itemize}
		\item $M_T$ is the set of sequences without repetition on $\mathcal X$.
		\item $s^{I_T} = \epsilon$.
		\item Interpret $\preceq$ as the prefix-order.
		\item We set $$f_\psi^{I_T}(\psi_1\dots\psi_n) = \begin{cases}
			\psi_1\dots\psi_n, &\text{if $\psi$ occurs in $[\psi_1]\dots[\psi_n]$,}\\
			\psi_1\dots\psi_n\psi, &\text{else.}			
		\end{cases}$$
		\item For propositional variables $P$, we set $$P^{I_T}\left(\psi_1\dots \psi_n\right) = P^I\left(g_{\psi_n}\left(\dots\left(g_{\psi_1}\left(s^I\right)\right)\dots\right)\right).$$
	\end{itemize}
\end{definition}
\begin{lemma}~\label{thm:prop-countermodel-reduction}
	Let $\mathcal M = (M, I)$ be a counter-model to $\mathcal \varphi^{CH}$.
	\begin{enumerate}
		\item $\psi$ is fulfilled at $f_{[\psi]}^I(u)$ for all $u\in M, \psi\in\mathcal X$.
		\item If $\psi$ is fulfilled at some $m\in M$ then $\psi$ is fulfilled at all $n\succeq^I m$.
	\end{enumerate}
\end{lemma}

@@

Now again due to Lemma~\ref{lemma:first-order-normal-form} it is sufficient to consider formulas $\varphi$ of the form $$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal X\wedge\bigwedge\mathcal Q\right)\to P$$ with flat clauses $\mathcal R$, implication clauses $\mathcal X$ and quantification clauses $\mathcal Q$, which after Herbrandization leaves us with $$\varphi^{CH} = \text{World}(s)\to K(\varphi)\to\left(\bigwedge R^{bS}\wedge \bigwedge X^{bS}\wedge\bigwedge Q^{bS}\right)\to P(s)$$
where $s$ is a new constant symbol, each formula in $\mathcal R^{bS}$ is of the form
$$\forall w \forall \vec x(s\preceq w\to\vec E(x, w)\to\bigwedge_i A_i(\vec a_i, u)\to\bigvee_j B_j(\vec b_j, u)),$$
each formula $\psi\in\mathcal X^{bS}$ is of the form
$$
	\resizebox{\hsize}{!}{$\forall w\forall\vec x(s\preceq w\to\vec E(\vec x, w)\to (w\preceq f_\psi(\vec x, w)\to A(\vec a, f_\psi(\vec x, w))\to B(\vec b, f_\psi(\vec x, w)))\to C(\vec c, w))$}
$$
and each formula $\psi\in\mathcal Q^{bS}$ is of the form
$$\forall w\forall\vec x(s\preceq w\to\vec E(\vec x, w)\to \forall y(E(y, f_\psi(\vec x, w))\to A(\vec a, f_\psi(\vec x, w)))\to B(\vec B, w))$$
with $f_\psi$ being a new function symbol.

As in the previous section we can define a tree counter-model for this.
\begin{definition}
	Suppose we are given a counter-model $\mathcal M = (M, I)$ for $\varphi^{CH}$. Define $\mathcal M_T = (M_T, I_T)$ as follows:
	\begin{itemize}
		\item $M_T$ are the sequences on $\{(\vec m, \psi)\:|\:\psi\in \mathcal F, \vec m\in M^n\}$.
		\item $r^{I_T} = s^{I_T} = \epsilon$.
		\item We interpret $\preceq$ as the prefix-order.
		\item We set $f_\psi^{I_T}(\vec m, (\vec m_1, \psi_1)\dots (\vec m_n, \psi_n)) = (\vec m_1, \psi_1)\dots (\vec m_n, \psi_n)(\vec m, \psi).$
		\item For predicate symbols $R$, we set $${R}^{I_T}(\vec m, (\vec m_1, \psi_1)\dots (\vec m_n, \psi_n)) = {R}^I(\vec m, f_{\psi_n}(\vec m_n, \dots(f_{\psi_1}(\vec m_1, s^I))\dots)).$$
	\end{itemize}
\end{definition}

@@

Even if we don't have the finiteness property of the propositional case, we have still managed to reduce the complexity of possible counter-models because we know that considering worlds of the form $$f_{\psi_1}(\vec m_1, f_{\psi_2}(\vec m_2, \dots f_{\psi_n}(\vec m_n, s)\dots))$$ is sufficient.
In particular, this allows us to completely eliminate $\preceq$, that is we remove the PartialOrder$(\preceq)$
from $K(\varphi)$ and adjust DomainSubset$(u, w)$ and Persistent$(u, w)$, e.g. remove quantification over $w$ and the precondition $u\preceq w$ from $K(\varphi)$ and replace DomainSubset$(u, w)$ with
$$
	\bigwedge\{\forall u\forall x\forall\vec y(E(x, u)\to E(x, f_\psi(\vec y, u)))\:|\:\psi\in\mathcal X\cup\mathcal Q\}
$$
and Persistent$(u, w)$ with
$$
	\bigwedge\{\forall\vec x\forall\vec y\forall u(A(\vec x, u)\to A(\vec x, f_\psi(\vec y, u)))\:|\:\psi\in\mathcal X\cup\mathcal Q, \text{$A$ a predicate in $\varphi$}\}.
$$Putting all of this together yields the following translation:
\begin{theorem}\label{thm:fo-translation}
	Let $\mathcal R,\mathcal X,\mathcal Q$ be as above. For every $n$-ary predicate $A$ define a new $n+1$-ary predicate of the same name. For every $\psi\in\mathcal \mathcal X\cup\mathcal Q$ define a new function symbol $f_\psi$. Consider a new binary predicate symbol $E$ and define $\vec E(\vec x, u) := \bigwedge\{E(x, u)\:|\:x\in\vec x\}$. Obtain $\mathcal R^\#$ by including
	$$
		\forall u\forall \vec x\left(\vec E(\vec x, u)\to\bigwedge_i A_i(\vec a_i, u)\to\bigvee_j B_j(\vec b_j, u)\right)
	$$
 	for each flat clause
 	$
 		\forall \vec x\left(\bigwedge_i A_i(\vec a_i)\to\bigvee_j B_j(\vec b_j)\right)\in\mathcal R,
 	$
 	$\mathcal X^\#$ by including
 	$$
 		\forall u\forall \vec x\left(A(\vec a, f_\psi(\vec x, u))\to B(\vec b, f_\psi(\vec x, u))\to C(\vec c, u)\right)
 	$$
 	for each implication clause
	$
		\psi = \forall \vec x\left(\left(A(\vec a)\to B(\vec b)\right)\to C(\vec c)\right)\in\mathcal X,
	$
	and $\mathcal Q^\#$ by including
	$$
		\forall u\forall \vec x\left(\forall y\left(A(\vec a, f_\psi(\vec x, u))\right)\to B(\vec b, u)\right)
	$$
	for each implication clause
	$
		\psi = \forall \vec x\left(\forall y\left(A(\vec a)\right)\to B(\vec b)\right)\in\mathcal X.
	$
	Then $\varphi$ is intuitionistically valid if and only if
	$$
		\varphi^\# := \exists xE(x, s)\to K(\varphi)\to \left(\bigwedge\mathcal R^\#\wedge\bigwedge\mathcal X^\#\wedge\bigwedge\mathcal Q^\#\right)\to P^\#(s)
	$$
	is classically valid, where $s$ is a new constant symbol and
	\begin{align*}
		K(\varphi) := &\:\bigwedge\left\{\forall\vec z\forall x\forall u((E(x, u)\to E(x, f_\psi(\vec z, u)))\:|\:\psi\in\mathcal X\cup\mathcal Q\right\}\wedge\\
		&\:\forall u\left(\exists xE(x, u)\to \bigwedge\{\forall\vec z(\vec E(\vec z, u)\to E(f(\vec z), u))\:|\:\text{$f$ a function in $\varphi$}\}\right)\wedge\\
		&\resizebox{.9\hsize}{!}{$\:\bigwedge\left\{\forall\vec z_1\forall \vec z_2\forall u(A(\vec z_1, u)\to A(\vec z_1, f_\psi(\vec z_2, u)))\:|\:\text{$\psi\in\mathcal X\cup\mathcal Q$, $A$ a predicate in $\varphi$}\right\}$.}
	\end{align*}
	The size of the obtained formula is linear in the input. However $|\mathcal X| + |\mathcal Q| + 1$ new function symbols and the new binary predicate $E$ were introduced and each $n$-ary predicate has been extended to a $n+1$-ary predicate.
\end{theorem}

>>>>

\begin{definition}
For $\psi\in\mathcal X$ let $[\psi]$ denote the corresponding clause in $\mathcal X^{b S}$. Suppose we are given a counter-model $\mathcal M = (M, I)$ for $\varphi^{CH}$.
We say $(A\to B)\to C\in\mathcal X$ is \emph{fulfilled} at $m\in M$ iff $A^I(m)\to B^I(m)$ is false or $C^I(m)$ is true. For $\psi\in\mathcal X$ define $$g_\psi : M\to M, m\mapsto\begin{cases}
		m,&\text{ if $\psi$ is fulfilled at $m$,}\\
		f^I_{\lbrack\psi\rbrack}(m),&\text{ else.}		
	\end{cases}$$Define a model $\mathcal M_T = (M_T, I_T)$ as follows:
	\begin{itemize}
		\item $M_T$ is the set of sequences without repetition on $\mathcal X$.
		\item $s^{I_T} = \epsilon$.
		\item Interpret $\preceq$ as the prefix-order.
		\item We set $$f_\psi^{I_T}(\psi_1\dots\psi_n) = \begin{cases}
			\psi_1\dots\psi_n, &\text{if $\psi$ occurs in $[\psi_1]\dots[\psi_n]$,}\\
			\psi_1\dots\psi_n\psi, &\text{else.}			
		\end{cases}$$
		\item For propositional variables $P$, we set $$P^{I_T}\left(\psi_1\dots \psi_n\right) = P^I\left(g_{\psi_n}\left(\dots\left(g_{\psi_1}\left(s^I\right)\right)\dots\right)\right).$$
	\end{itemize}
\end{definition}
\begin{lemma}~\label{thm:prop-countermodel-reduction}
	Let $\mathcal M = (M, I)$ be a counter-model to $\mathcal \varphi^{CH}$.
	\begin{enumerate}
		\item $\psi$ is fulfilled at $f_{[\psi]}^I(m)$ for all $m\in M, \psi\in\mathcal X$.
		\item If $\psi$ is fulfilled at some $m\in M$ then $\psi$ is fulfilled at all $n\succeq^I m$.
	\end{enumerate}
\end{lemma}

@@

Now again due to Lemma~\ref{lemma:first-order-normal-form} it is sufficient to consider formulas $\varphi$ of the form $$\left(\bigwedge\mathcal R\wedge\bigwedge\mathcal X\wedge\bigwedge\mathcal Q\right)\to P$$ with flat clauses $\mathcal R$, implication clauses $\mathcal X$ and quantification clauses $\mathcal Q$, which after Herbrandization leaves us with $$\varphi^{CH} = K(\varphi)\to\text{World}(s)\to\vec E(\vec r, s) \to \left(\bigwedge R^{bS}\wedge \bigwedge X^{bS}\wedge\bigwedge Q^{bS}\right)\to P(s)$$
where $\vec r, s$ are new constant symbols, each formula in $\mathcal R^{bS}$ is of the form
$$\forall w \forall \vec x(s\preceq w\to\vec E(x, w)\to\bigwedge_i A_i(\vec a_i, u)\to\bigvee_j B_j(\vec b_j, u)),$$
each formula $\psi\in\mathcal X^{bS}$ is of the form
$$
	\resizebox{\hsize}{!}{$\forall w\forall\vec x(s\preceq w\to\vec E(\vec x, w)\to (w\preceq f_\psi(\vec x, w)\to A(\vec a, f_\psi(\vec x, w))\to B(\vec b, f_\psi(\vec x, w)))\to C(\vec c, w))$}
$$
and each formula $\psi\in\mathcal Q^{bS}$ is of the form
$$\forall w\forall\vec x(s\preceq w\to\vec E(\vec x, w)\to \forall y(E(y, f_\psi(\vec x, w))\to A(\vec a, f_\psi(\vec x, w)))\to B(\vec B, w))$$
with $f_\psi$ being a new function symbol.

As in the previous section we can define a tree counter-model for this.
\begin{definition}
	Suppose we are given a counter-model $\mathcal M = (M, I)$ for $\varphi^{CH}$. W.l.o.g. suppose $\{m\in M\:|\:\text{World}(m)\}$ has a minimal element, say $w_0$. Define $\mathcal M_T = (M_T, I_T)$ as follows:
	\begin{itemize}
		\item $M_T$ is the union of $M$ and sequences $(\vec m_1, \psi_1)\dots (\vec m_n, \psi_n)$ such that $\vec E^I(\vec m_1, w_0)$ and $\vec E^I(\vec m_i, f_{\psi_i}(\vec m_i, f_{\psi_{i-1}}^I(\vec m_{i-1}, \dots f_{\psi_1}^I(\vec m_1, w_0)\dots)))$ for $i > 0$ hold.
		\item $s^{I_T} = w_0$, $\vec r^{I_T} = \vec r^I$.
		\item We interpret $\preceq$ trivially on $M$ and as the prefix-order elsewhere.
		\item Have $E^{I_T}(m, (\vec m_1, \psi_1)\dots (\vec m_n, \psi_n))$ iff $E^I(m, f_{\psi_n}^I(\vec m_n, \dots f_{\psi_1}^I(\vec m_1, w_0)\dots))$.
		\item We set $f_\psi^{I_T}(\vec m, (\vec m_1, \psi_1)\dots (\vec m_n, \psi_n)) = (\vec m_1, \psi_1)\dots (\vec m_n, \psi_n)(\vec m, \psi)$ and arbitrary else.
		\item For predicate symbols $R$, we set $${R}^{I_T}(\vec m, (\vec m_1, \psi_1)\dots (\vec m_n, \psi_n)) = {R}^I(\vec m, f_{\psi_n}^I(\vec m_n, \dots(f_{\psi_1}^I(\vec m_1, s^I))\dots))$$ and arbitrary otherwise, for function symbols $f^{I_T}(\vec m) = f^I(\vec m)$ and arbitrary else.
	\end{itemize}
\end{definition}

@@

Even if we don't have the finiteness property of the propositional case, we have still managed to reduce the complexity of possible counter-models because we know that considering worlds of the form $$f_{\psi_1}(\vec m_1, f_{\psi_2}(\vec m_2, \dots f_{\psi_n}(\vec m_n, s)\dots))$$ is sufficient.
In particular, this allows us to completely eliminate $\preceq$, that is we remove the PartialOrder$(\preceq)$
from $K(\varphi)$ and adjust DomainSubset$(u, w)$ and Persistent$(u, w)$, e.g. remove quantification over $w$ and the precondition $u\preceq w$ from $K(\varphi)$ and replace DomainSubset$(u, w)$ with
$$
	\bigwedge\{\forall u\forall x\forall\vec y(E(x, u)\to E(x, f_\psi(\vec y, u)))\:|\:\psi\in\mathcal X\cup\mathcal Q\}
$$
and Persistent$(u, w)$ with
$$
	\bigwedge\{\forall\vec x\forall\vec y\forall u(A(\vec x, u)\to A(\vec x, f_\psi(\vec y, u)))\:|\:\psi\in\mathcal X\cup\mathcal Q, \text{$A$ a predicate in $\varphi$}\}.
$$Putting all of this together yields the following translation:
\begin{theorem}\label{thm:fo-translation}
	Let $\mathcal R,\mathcal X,\mathcal Q$ be as above. For every $n$-ary predicate $A$ define a new $n+1$-ary predicate of the same name. For every $\psi\in\mathcal \mathcal X\cup\mathcal Q$ define a new function symbol $f_\psi$. Consider a new binary predicate symbol $E$ and define $\vec E(\vec x, u) := \bigwedge\{E(x, u)\:|\:x\in\vec x\}$. Obtain $\mathcal R^\#$ by including
	$$
		\forall u\forall \vec x\left(\vec E(\vec x, u)\to\bigwedge_i A_i(\vec a_i, u)\to\bigvee_j B_j(\vec b_j, u)\right)
	$$
 	for each flat clause
 	$
 		\forall \vec x\left(\bigwedge_i A_i(\vec a_i)\to\bigvee_j B_j(\vec b_j)\right)\in\mathcal R,
 	$
 	$\mathcal X^\#$ by including
 	$$
 		\forall u\forall \vec x\left(A(\vec a, f_\psi(\vec x, u))\to B(\vec b, f_\psi(\vec x, u))\to C(\vec c, u)\right)
 	$$
 	for each implication clause
	$
		\psi = \forall \vec x\left(\left(A(\vec a)\to B(\vec b)\right)\to C(\vec c)\right)\in\mathcal X,
	$
	and $\mathcal Q^\#$ by including
	$$
		\forall u\forall \vec x\left(\forall y\left(A(\vec a, f_\psi(\vec x, u))\right)\to B(\vec b, u)\right)
	$$
	for each implication clause
	$
		\psi = \forall \vec x\left(\forall y\left(A(\vec a)\right)\to B(\vec b)\right)\in\mathcal X.
	$
	Then $\varphi$ is intuitionistically valid if and only if
	$$
		\varphi^\# := K(\varphi)\to\text{World}(s)\to\vec E(\vec r, s) \to \left(\bigwedge\mathcal R^\#\wedge\bigwedge\mathcal X^\#\wedge\bigwedge\mathcal Q^\#\right)\to P^\#(s)
	$$
	is classically valid, where $s, \vec r$ are new constants introduced by Herbrandization of free variables,
	\begin{align*}
		K(\varphi) := &\:\bigwedge\left\{\forall\vec z\forall x\forall u((E(x, u)\to E(x, f_\psi(\vec z, u)))\:|\:\psi\in\mathcal X\cup\mathcal Q\right\}\wedge\\
		&\:\forall u\left(\exists xE(x, u)\to \bigwedge\{\forall\vec z(\vec E(\vec z, u)\to E(f(\vec z), u))\:|\:\text{$f$ a function in $\varphi$}\}\right)\wedge\\
		&\resizebox{.9\hsize}{!}{$\:\bigwedge\left\{\forall\vec z_1\forall \vec z_2\forall u(A(\vec z_1, u)\to A(\vec z_1, f_\psi(\vec z_2, u)))\:|\:\text{$\psi\in\mathcal X\cup\mathcal Q$, $A$ a predicate in $\varphi$}\right\}$.}
	\end{align*}
	The size of the obtained formula is linear in the input. However $|\mathcal X| + |\mathcal Q| + 1$ new function symbols and the new binary predicate $E$ were introduced and each $n$-ary predicate has been extended to a $n+1$-ary predicate.
\end{theorem}
